
## R Markdown

## Including Plots

You can also embed plots, for example:



Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

The Research Questions had been modified based on the comments given for the previous assignments as below:

1. What is the trend of all the five-leading cause of deaths over time for each year?

2. What is the maximum and minimum number of expected deaths, observed deaths, potentially excess deaths for a particular cause of death in each age group?

3. What is the trend of the deaths over various regions of US? Which region has the maximum and minimum number of deaths?

4. What is the trend of expected deaths, observed deaths, potentially excess deaths in each region for each benchmark in each locality?

5. What is the ratio between the observed deaths and the Population?What is the trend of the ratios over time by region?

6. What is the ratio between the expected deaths and the Population?What is the trend of the ratios over time by region?

7. Are the ratios in questions 5 and 6 correlated?

### This document consists of R script which will help in providing answers to the above questions through R plots.

NCHS_New1 <- read.csv("NCHS.csv")

str(NCHS_New1)

summary(NCHS_New1)

sum(is.na(NCHS_New1))

aft_mis_NCHS_New1<-na.omit(NCHS_New1)

sum(is.na(aft_mis_NCHS_New1))

str(aft_mis_NCHS_New1)

duplicated(NCHS_New1)

str(NCHS_New1)

aft_mis_NCHS_New2<- aft_mis_NCHS_New1[!(aft_mis_NCHS_New1$State == "United States"),]

dat <- aft_mis_NCHS_New2

head(dat)

library(noncensus)

data("states")

states

Reg <- states[,c(1,3)]

Reg

Health_data <- merge(dat,Reg,by.x = "State.FIPS.Code",by.y = "state")

head(Health_data)

### write.csv(Health_data,"Health data.csv")

# 1.trend of all the five leading cause of Observed deaths over time for each region

Data_all <- Health_data[,c(2:3,9,14)]

head(Data_all)

library(dplyr)

Data_allNew <- Data_all  %>%
  group_by(region, Year, Cause.of.Death) %>%
  summarise_all(funs(mean))

Data_allNew

library(ggplot2)

ggplot(data = Data_allNew, aes(x=Year, y = Observed.Deaths, colour = region, group = region)) + labs(title="Trend of Observed Deaths by Year", x="Year",y="Observed Deaths") +
  geom_line()+scale_x_continuous(breaks=seq(2005, 2015, 3)) + facet_wrap(~Cause.of.Death,scales = "free_y")
  
```{r Data_allNew, echo=FALSE}
ggplot(data = Data_allNew, aes(x=Year, y = Observed.Deaths, colour = region, group = region)) + labs(title="Trend of Observed Deaths by Year", x="Year",y="Observed Deaths") +
  geom_line()+scale_x_continuous(breaks=seq(2005, 2015, 3)) + facet_wrap(~Cause.of.Death,scales = "free_y")
```


# 2. Max and Min value of all types of deaths in Each age group

Data_sec <- Health_data[,c(3,6,9,11,12)]

str(Data_sec)

library(dplyr)

Data_sec_New <- Data_sec  %>%
  group_by(Age.Range, Cause.of.Death) %>%
  summarise_all(funs(mean))

Data_sec_New

library(tidyr)

long_data_sec_New <- gather(Data_sec_New,"Death_Type", "Deaths", Observed.Deaths, Expected.Deaths, Potentially.Excess.Deaths)

long_data_sec_New

library(ggplot2)

ggplot(data = long_data_sec_New, aes(x=factor(Age.Range), y = Deaths, colour = Death_Type, group = Death_Type)) + labs(title="Trend of Deaths by Age Group", x="Age Group",y=" Number of Deaths") + 
  geom_line() + facet_wrap(~Cause.of.Death,scales = "free")
  
```{r long_data_sec_New, echo=FALSE}
ggplot(data = long_data_sec_New, aes(x=factor(Age.Range), y = Deaths, colour = Death_Type, group = Death_Type)) + 
  geom_line() + facet_wrap(~Cause.of.Death,scales = "free")
```


# 3. Max and Min Deaths by region

library(dplyr)

Data_region <- Health_data[,c(9,11,12,14)] %>%
  group_by(region) %>%
  summarise_all(funs(mean))

Data_region

library(tidyr)

long_data_region <- gather(Data_region,"Death_Type", "Avg_Deaths", Observed.Deaths, Expected.Deaths, Potentially.Excess.Deaths)

long_data_region

library(ggplot2)

ggplot(long_data_region,aes(x = region,y = Avg_Deaths,fill = Death_Type)) + labs(title="Trend of Average of Deaths by Region", x="Region",y="Mean of Deaths") + geom_bar(position = "dodge", stat = "identity")

```{r long_data_region, echo=FALSE}
ggplot(long_data_region,aes(x = region,y = Avg_Deaths,fill = Death_Type)) + labs(title="Trend of Average of Deaths by Region", x="Region",y="Mean of Deaths") + 
  geom_bar(position = "dodge", stat = "identity")
```

# 4. Trend of Deaths by locality

library(dplyr)

Data_Locality <- Health_data[,c(8,9,11,12)] %>%
  group_by(Locality) %>%
  summarise_all(funs(mean))

Data_Locality

library(tidyr)

long_Data_Locality <- gather(Data_Locality,"Death_Type", "Avg_Deaths", Observed.Deaths, Expected.Deaths, Potentially.Excess.Deaths)

long_Data_Locality

library(ggplot2)

ggplot(long_Data_Locality,aes(x = Locality,y = Avg_Deaths,fill = Death_Type)) + labs(title="Trend of Average of Deaths by Locality", x="Region",y="Mean of Deaths") + geom_bar(position = "dodge", stat = "identity")

```{r long_Data_Locality, echo=FALSE}
ggplot(long_Data_Locality,aes(x = Locality,y = Avg_Deaths,fill = Death_Type)) + labs(title="Trend of Average of Deaths by Locality", x="Region",y="Mean of Deaths") + geom_bar(position = "dodge", stat = "identity")
```

# 5. Ratio of observed deaths and population over various regions

Health_data$percent_obs.deaths <- Health_data$Observed.Deaths / Health_data$Population

head(Health_data)

Obs.Deaths_region <- Health_data[,c(14:15)] %>%
  group_by(region) %>%
  summarise_all(funs(mean))

Obs.Deaths_region

library(ggplot2)

ggplot(Obs.Deaths_region,aes(x = region,y = percent_obs.deaths,fill = region)) + labs(title="Trend of Observed Deaths percentage", x="Region",y="Observed Deaths/Population") + geom_bar(stat = "identity")

```{r Obs.Deaths_region, echo=FALSE}
ggplot(Obs.Deaths_region,aes(x = region,y = percent_obs.deaths,fill = region)) + labs(title="Trend of Observed Deaths percentage", x="Region",y="Observed Deaths/Population") + geom_bar(stat = "identity")
```

# 6. Ratio of Expected Deaths to Population over regions

Health_data$percent_exp.deaths <- Health_data$Expected.Deaths / Health_data$Population

head(Health_data)

exp.Deaths_region <- Health_data[,c(14,16)] %>%
  group_by(region) %>%
  summarise_all(funs(mean))

exp.Deaths_region

library(ggplot2)

ggplot(exp.Deaths_region,aes(x = region,y = percent_exp.deaths,fill = region)) + labs(title="Trend of Expected Deaths percentage", x="Region",y="Expected Deaths/Population")+ geom_bar(stat = "identity")

```{r exp.Deaths_region, echo=FALSE}
ggplot(exp.Deaths_region,aes(x = region,y = percent_obs.deaths,fill = region)) + labs(title="Trend of Observed Deaths percentage", x="Region",y="Observed Deaths/Population") + geom_bar(stat = "identity")
```

# 7. Correlation between both the ratios in Questions 5 and 6

cor(Health_data$percent_obs.deaths,Health_data$percent_exp.deaths)

#### There is a Positive correlation of 0.9469511 between the two ratios
There is a 0.94 correlation between the two ratios which means that the ratios are in sync for 94% of the records. But the ratios recorded a non-corrleation i.e one ratio increased and the other ratio decreased in 6% of the cases. 
